\documentclass[11pt]{article}
\newcommand{\tb}{\textbf}
\usepackage{enumerate}
\usepackage{amsmath}

\begin{document}
	\title{\textbf{Economic Analysis - Engineering}}
	\maketitle
	\section{Revenue, Cost, Profit}
	\begin{enumerate}
		\item $Profit = Revenue - Cost$ 
	\end{enumerate} 
	\subsection{Example 1}
	\indent Ace Manufacturer is a major player in the lawn sprinkler business. To produce these sprinklers Acme's fixed cost is \$55,000 per most for the production. The variable cost for each sprinkler is \$15.50 per unit. The selling price for these sprinklers is fixed at \$60 per unit. Acme has the capacity to produce 3,000 units per month. 
	\begin{enumerate}
		\item What is the maximum profit Acme can expect at full capacity?
		\item If acme produces 1000 sprinklers per month, will the company experience a profit or loss?
	\end{enumerate}
\paragraph
1. 
\\
 $x=3000$
\\
$Profit = Revenue - Cost$
\\
$\{(\frac{3000 sprink}{month})(\frac{60}{sprink}) - \frac{55000}{month} - (\frac{3000 sprink}){month})(\frac{15.50}{sprink})\} = \$78,500$
\\
\\
2. $x=1000$ 
\\
$Profit = (1000)(60) - 55000 -(1000)(15.50) = -10,500$	
	
\subsection{Problem 1}
Scoot-Yer-Boot produces parts for electric scooters.  One of the parts they produce is the fork (connects the handle to the front wheel).  Consider the following information about the production process:
\\
Fixed production costs 	   \textbf{\$53,000 per month}
\\
Metal to make one fork 	   \textbf{\$18 per fork}
\\
Number of forks made each day 	   \textbf{870 forks per day}
\\
Percent of forks that pass quality inspection (are usable) 		\textbf{90\%}
\\
Selling price of usable forks 	  \textbf{\$32 per fork}
\begin{enumerate}
	\item Assuming there are 20 Production days per month, what is the expected monthly profit?
\end{enumerate}
1. 
\\
$ Profit = (\frac{\$32}{fork})(\frac{870 forks}{day})(\frac{20 days}{month})(\frac{9}{10}) - (\frac{\$18}{fork})(\frac{870 forks}{day})(\frac{20 days}{month}) - (\frac{\$53,000}{month})$
\\
\subsection{Problem 2}
The owners of a mine are trying to estimate the profitability of a mine containing \textbf{100,000 tons of ore}. The ore recovery method being considered will result in the removal of \textbf{61,000 tons of ore} and will cost \textbf{\$21 per ton of ore removed}. Subsequent processing of the removed ore results in \textbf{300 lbs of metal per ton of ore} processed. The cost of this processing is \textbf{\$45 per ton of ore} processed. The recovered metal can be sold for \textbf{\$0.83 per lb.}
\\
1. 
What is the expected total profit from the mine?
\\
$ Profit = (61,000 tons)*(3/10)*(\frac{2000 lb}{ton})(\$0.83/lb) - (61000 tons)(\$21/ton) - (61,000 tons)*(\$45/ton)$

\end{document}
